---
title: "2020 HINTS"
output: html_document
date: "2026-03-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (dplyr) # open 'dplyr' package
library(skimr) # open 'skimr' package
read.csv(hints.data)
```

# 2020 Health Information National Trends Survey (HINTS) Analysis

## Statistical Descriptive Analysis

### Age

The average age HINTS participant was approximately 56, the youngest participant was 18 and the oldest participant was 100 years of age. Furthermore, the median age was 58.

```{r}
summary (hints.data$Age, na.rm=TRUE)
```

Statistical summary of age based on three different smoking groups. The average age for current smokers is 54, for former smokers is 61, and for individuals who never smoked is 53.

```{r}
hints.data %>%
  group_by (smokeStat) %>%
  summarise (
    mean_age = mean(Age, na.rm=TRUE), # displays mean of 'Age'
    median_age = median(Age, na.rm=TRUE), # displays median of 'Age'
    sd_age = sd(Age, na.rm=TRUE), # displays standard deviation of 'Age'
    q25_age = quantile(Age, 0.25, na.rm=TRUE), # displays 0.25 qt of 'Age'
    q75_age = quantile(Age, 0.75, na.rm=TRUE), # displays 0.75 qt of 'Age' 
    var_age = var(Age, na.rm = TRUE)
  )
```

Statistical summary of age utilizing the 'skimr' package.

```{r echo=TRUE}
skim(hints.data$Age) # provides statistical summary for 'Age' using 'skim' package
```

### Average Drinks Per Week

The average drinks per week HINTS participant drank was about 3, the lowest was 0, and the highest was 70. 

```{r}
summary(hints.data$AvgDrinksPerWeek, na.rm = TRUE) # Descriptive summary 'AvgDrinksPerWeek' variable
```

Statistical summary of average drinks per week based on three different smoking groups. THe average drinks per week for current smokers is 5, for former smokers is 3, and for people who never smoked is 1. 

```{r}
hints.data %>%
  group_by(smokeStat) %>%
  summarise(
    mean_avgdrinksperweek = mean(AvgDrinksPerWeek, na.rm = TRUE), # Mean
    median_avgdrinksperweek = median(AvgDrinksPerWeek, na.rm = TRUE), # Median 
    sd_avgdrinksperweek = sd(AvgDrinksPerWeek, na.rm = TRUE), # Standard deviation
    q25_avgdrinksperweek = quantile(AvgDrinksPerWeek, 0.25, na.rm = TRUE), # 25th percentile
    q75_avgdrinksperweek = quantile(AvgDrinksPerWeek, 0.75, na.rm = TRUE) # 75th percentile
    var_avgdrinksperweek = var(AvgDrinksPerWeek, na.rm = TRUE) # Variance
  )
```

### Weekly Minutes Moderate Exercise

### BMI

## Univariate Visualizations

### Quality of Care

### Race and Ethnicity

### Age

The histogram is displaying the age distribution of HINTS participants. The age distribution is left-skewed indicating that most survey participants were older.

```{r}
hist(hints.data$Age, main = "histogram of age")
```

### BMI

The boxplot is providing the five-number summary of Body Mass Index (BMI) of survey participants. The average BMI of HINTS participants was 28.61. The maximum was 73.80 and the minimum was 10.90.

```{r}
boxplot (hints.data$BMI, main = "BMI Boxplot")
```

```{r}
summary(hints.data$BMI) # provide statistical summary of 'BMI' 
```
