---
title: "2020 HINTS"
output: html_document
date: "2026-03-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (dplyr) # open 'dplyr' package
library(skimr) # open 'skimr' package
```

# 2020 Health Information National Trends Survey (HINTS) Analysis

## Statistical Descriptive Analysis

### Age

The average age HINTS participant was approximately 56, the youngest participant was 18 and the oldest participant was 100 years of age. Furthermore, the median age was 58.

```{r}
summary (hints.data$Age, na.rm=TRUE)
```

Statistical summary of age based on three different smoking groups.

```{r}
hints.data %>%
  group_by (smokeStat) %>%
  summarise (
    mean_age = mean(Age, na.rm=TRUE), # displays mean of 'Age'
    median_age = median(Age, na.rm=TRUE), # displays median of 'Age'
    sd_age = sd(Age, na.rm=TRUE), # displays standard deviation of 'Age'
    q25_age = quantile(Age, 0.25, na.rm=TRUE), # displays 0.25 qt of 'Age'
    q75_age = quantile(Age, 0.75, na.rm=TRUE), # displays 0.75 qt of 'Age' 
    var_age = var(Age, na.rm = TRUE)
  )
```

Statistical summary of age utilizing the 'skimr' package.

```{r}
skim(hints.data$Age) # provides statistical summary for 'Age' using 'skim' package
```

### Average Drinks Per Week

### Weekly Minutes Moderate Exercise

### BMI

## Univariate Visualizations

### Quality of Care

### Race and Ethnicity

### Age

The histogram is displaying the age distribution of HINTS participants.

```{r}
hist(hints.data$Age, main = "histogram of age")
```

### BMI

The Boxplot is providing the five-number summary of Body Mass Index (BMI) of survey participants.

```{r}
boxplot (hints.data$BMI, main = "BMI Boxplot")
```
