---
title: "2020 HINTS Project"
author: "Katirna Alatorre, Katrina Huynh, Jaime Rosas, and Linda Zavala"
output: html_document
date: "2026-03-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (dplyr) # open 'dplyr' package
library(skimr) # open 'skimr' package
```

# 2020 Health Information National Trends Survey (HINTS) Analysis

Project Objectives: This project analyzes data from the 2020 Health Information National Trends Survey (HINTS). The goal is to examine health behaviors and demographic characteristics.

## HINTS Data Import

```{r}
hints.data <- read.csv("C:/Users/zlind/Downloads/HINTSData_2020_clean.csv")
```

The following data focuses on the variables:

Age, AvgDrinksPerWeek, WeeklyMinutesModerateExcersice, BMI, QualityCare, BirthGender, smokeStat, and RaceEthn5. 

## Statistical Descriptive Analysis

### Age

The average age HINTS participant was approximately 56, the youngest participant was 18 and the oldest participant was 100 years of age. Furthermore, the median age was 58.

```{r}
summary(hints.data$Age, na.rm=TRUE)
```

### Average Drinks Per Week

### Weekly Minutes Moderate Exercise

### BMI

The lowest BMI value among the HINTS participants was 10.90 and the highest was approximately 73.80. 
The average BMI among participants was 28.61. 

```{r}
summary(hints.data$BMI, na.rm = TRUE)
```

```{r}
var(hints.data$BMI, na.rm = TRUE)
```

### Function for Descriptive Statistics 
```{r}
#creating a function to summarize the statisical summaries of the continous variables of intrest  

my_summary_func <- function(x) { #defines a function
  list(   
    Mean = mean(x),#mean
    SD = sd (x), #standard deviation
    Variance = var(x), #vairance
    Five.Num.Summary = summary(x) #five summary
  )
}

#applying my_summary_function to all continous variables
my_summary_func(hints.data$Age)

my_summary_func(hints.data$BMI)

my_summary_func(hints.data$AvgDrinksPerWeek)

my_summary_func(hints.data$WeeklyMinutesModerateExercise)
```

## Univariate Visualizations

### Quality of Care

```{r}
#create bar graph for quality of care distribtion 
barplot(table(hints.data$QualityCare), #create table
        main = "Distribution of Quality of Care", #names graph
        xlab = "Quality of Care", #labels x-axis
        ylab = "Frequency") # labels y-axis
```

The bar chart displays how participants perceived the quality of healthcare from a ranking of 1-5 (the numbers correspond to ordered response categories).The majority of participants chose 'very good'. 

Key:

  1 Excellent
  
  2 Very Good
  
  3 Good 
  
  4 Fair
  
  5 Poor

### Race and Ethnicity

```{r}
#create bar graph for race and ethnicty
barplot(table (hints.data$RaceEthn5),
        main = "Distribution of Race and Ethnicity", #names graph
        xlab = "Race", #labels x-axis
        ylab = "Frequency") #labels y-axis
```

The bar chart represents the distribution of the Race and Ethnicity of the participants. Each number represents a different option. The majority of the participants were non-Hispanic white.

Key:

  1 Non-Hispanic White
  
  2 Non-Hispanic Black or African
  American
  
  3 Hispanic
  
  4 Non-Hispanic Asian
  
  5 Non-Hispanic Other
  
### Age

### BMI

```{r}
boxplot (hints.data$BMI, main = "BMI Boxplot")
```
